### 2024-04-29 thesis/input_files/metrics.tex
correct names in contents,list of fig and tables

Proper evaluation methods guide researchers in choosing the model that best fits their needs; thus, this chapter is dedicated to the most common evaluation metrics adopted by academics in the field of EF. Error metrics and measures vary depending on wethter we are concerned with point or probabilistic forecasts. Additionally, note that the latter can take different forms which therefore requires different measures.
\section{MAE}\label{mae}
Consider the time series with actual values given by $\mathrm{L}=(L_{n+1}, L_{n+2},\dots, L_{n+h})$
and its h step ahead point forecast $\mathrm{\hat{L}}=(\hat{L}_{n+1}, \hat{L}_{n+2},\dots, \hat{L}_{n+h})$ the mean absolute error is defined as
\begin{definition}
    $\mathrm{MAE}(\mathrm{L},\mathrm{\hat{L}})=\frac{1}{h}\|\mathrm{L}-\mathrm{\hat{L}}\|_{1}=\frac{1}{h}\sum\limits_{k=1}^{h}|L_{n+k}-\hat{L}_{n+k}|$
\end{definition}

\section{RMSE}\label{rmse}
\begin{definition}
    $\mathrm{RMSE(\mathrm{L}, \mathrm{\hat{L}})}=\frac{1}{\sqrt{h}}\|\mathrm{L}-\mathrm{\hat{L}}\|_{2}=\sqrt{\frac{\sum\limits_{k=1}^{h}(L_{n+k}- \hat{L}_{n+k})^2}{h}}$
\end{definition}

MAE and RMSE posses the useful property of being expressed in the same units of the data thus enabling meaningful comparisons.
However, a drawback of such measures is that we cannot use them to compare accuracy between time series wich have different magnitudes. For instance, a day ahead error of 1kWh is negligible when considering a daily demand of 100kWn while the same error is considerably big when daily demand is 2kWh. This consideration leads to relative accuracy scores, between those the MAPE is by far the most popular.

\section{MAPE}\label{mape}. 
\begin{definition}
    $\mathrm{MAPE}(\mathrm{L},\mathrm{\hat{L}})=\frac{100}{h}\sum\limits_{k=1}^{h}\frac{|L_{n+k}-\hat{L}_{n+k}|}{|L_{n+k}|}$
\end{definition}

\section{RMSPE}\label{rmspe}
\begin{definition}
    $\mathrm{RMSPE}(\mathrm{L},\mathrm{\hat{L}})=100\cdot\sqrt{\frac{1}{h}\sum\limits_{k=1}^{h} \left(\frac{|L_{n+k}-\hat{L}_{n+k}|}{|L_{n+k}|}\right)^2}$
\end{definition}
Note, MAPE and RMSPE may not be appropriate for series which have zero or very small values, for example, electricity demand at the household level; the result is a large score regardless of the absolute errors.
Scaled errors constitute a robust family of scores.
\section{MASE}\label{mase}
\begin{definition}
    $\mathrm{MASE}(\mathrm{L},\mathrm{\hat{L}})=\frac{1}{h}\sum\limits_{k=1}^N\frac{|L_{n+k}-\hat{L}_{n+k}|}{\frac{1}{h-1}\sum\limits_{k=2}^{h}|L_{k}-L_{k-1}|}$
\end{definition}
In the denominator we have the error of the naïve/persistence model. 
In this model, the current demand makes up the prediction for the next time step; that is $\mathrm{\hat{L}^{\mathrm{naive}}}_{n+1}=\mathrm{L}_{n}$.
\section{RMSSE}\label{rmsse}
\begin{definition}
    $\mathrm{RMSSE}(\mathrm{L},\mathrm{\hat{L}})=\sqrt{\frac{1}{h}\sum\limits_{k=1}^N\left(\frac{|L_{n+k}-\hat{L}_{n+k}|}{\frac{1}{h-1}\sum\limits_{k=2}^{h}|L_{k}-L_{k-1}|}\right)^2}$
\end{definition}

@@

    \caption{Pinball loss with q=0.2 \cite{haben2023core}}

@@

Where delta is the PI width, that is $\delta=Upp_t-Low_t$. This score penalizes observations falling outside the PI and rewards narrow PI.
\section{CRPS}
The continous ranked probability score measures the differece between the estimated cumulative distribution $\hat{F}$ and the empirical Cdf.

@@

\section{PIT}
The probability integral transform is a method to assess visually the quality of probabilistic forecasts. PIT is obtained by applying a cdf $F$ to your data; if applying such cdf to the data results in a uniform distributed PIT then $F$ is a valid prediction. If not, $F$ is not the suited cdf for the considered data. Figure \ref{fig:pit} provides an example, applying the true cdf results in a well calibrated PIT (left). Alternatively, applying a bad cdf results in either a overdispersed (middle) or underdispersed (right) PIT.

>>>>

Proper evaluation methods guide researchers in choosing the model that best fits their needs; thus, this chapter is dedicated to the most common evaluation metrics adopted by academics in the field of EF. Error metrics and measures vary depending on wethter we are concerned with point or probabilistic forecasts. Additionally, note that the latter can take different forms which therefore requires different measures.
\section{Mean absolute error}\label{mae}
Consider the time series with actual values given by $\mathrm{L}=(L_{n+1}, L_{n+2},\dots, L_{n+h})$
and its h step ahead point forecast $\mathrm{\hat{L}}=(\hat{L}_{n+1}, \hat{L}_{n+2},\dots, \hat{L}_{n+h})$ the mean absolute error (MAE) is defined as
\begin{definition}
    $\mathrm{MAE}(\mathrm{L},\mathrm{\hat{L}})=\frac{1}{h}\|\mathrm{L}-\mathrm{\hat{L}}\|_{1}=\frac{1}{h}\sum\limits_{k=1}^{h}|L_{n+k}-\hat{L}_{n+k}|$
\end{definition}

\section{Root mean squared error}\label{rmse}
\begin{definition}
    $\mathrm{RMSE(\mathrm{L}, \mathrm{\hat{L}})}=\frac{1}{\sqrt{h}}\|\mathrm{L}-\mathrm{\hat{L}}\|_{2}=\sqrt{\frac{\sum\limits_{k=1}^{h}(L_{n+k}- \hat{L}_{n+k})^2}{h}}$
\end{definition}

MAE and root mean squared error (RMSE) posses the useful property of being expressed in the same units of the data thus enabling meaningful comparisons.
However, a drawback of such measures is that we cannot use them to compare accuracy between time series wich have different magnitudes. For instance, a day ahead error of 1kWh is negligible when considering a daily demand of 100kWn while the same error is considerably big when daily demand is 2kWh. This consideration leads to relative accuracy scores, between those the mean absolute percentage error (MAPE) is by far the most popular.

\section{Mean absolute percentage error}\label{mape}. 
\begin{definition}
    $\mathrm{MAPE}(\mathrm{L},\mathrm{\hat{L}})=\frac{100}{h}\sum\limits_{k=1}^{h}\frac{|L_{n+k}-\hat{L}_{n+k}|}{|L_{n+k}|}$
\end{definition}

\section{Root mean squared percentage error}\label{rmspe}
\begin{definition}
    $\mathrm{RMSPE}(\mathrm{L},\mathrm{\hat{L}})=100\cdot\sqrt{\frac{1}{h}\sum\limits_{k=1}^{h} \left(\frac{|L_{n+k}-\hat{L}_{n+k}|}{|L_{n+k}|}\right)^2}$
\end{definition}
Note, MAPE and root mean squared percentage error (RMSPE) may not be appropriate for series which have zero or very small values, for example, electricity demand at the household level; the result is a large score regardless of the absolute errors.
Scaled errors constitute a robust family of scores.
\section{Mean absolute scaled error}\label{mase}
\begin{definition}
    $\mathrm{MASE}(\mathrm{L},\mathrm{\hat{L}})=\frac{1}{h}\sum\limits_{k=1}^N\frac{|L_{n+k}-\hat{L}_{n+k}|}{\frac{1}{h-1}\sum\limits_{k=2}^{h}|L_{k}-L_{k-1}|}$
\end{definition}
In the denominator we have the error of the naïve/persistence model. 
In this model, the current demand makes up the prediction for the next time step; that is $\mathrm{\hat{L}^{\mathrm{naive}}}_{n+1}=\mathrm{L}_{n}$.
\section{Root mean squared scaled error}\label{rmsse}
\begin{definition}
    $\mathrm{RMSSE}(\mathrm{L},\mathrm{\hat{L}})=\sqrt{\frac{1}{h}\sum\limits_{k=1}^N\left(\frac{|L_{n+k}-\hat{L}_{n+k}|}{\frac{1}{h-1}\sum\limits_{k=2}^{h}|L_{k}-L_{k-1}|}\right)^2}$
\end{definition}

@@

    \caption{Pinball loss with $q=0.2$ \cite{haben2023core}}

@@

Where delta is the prediction interval (PI) width, that is $\delta=Upp_t-Low_t$. This score penalizes observations falling outside the PI and rewards narrow PI.
\section{Continous ranked probability score}
The continous ranked probability score (CRPS) measures the differece between the estimated cumulative distribution $\hat{F}$ and the empirical Cdf.

@@

\section{Probability integral transform}
The probability integral transform (PIT) is a method to assess visually the quality of probabilistic forecasts. PIT is obtained by applying a cdf $F$ to your data; if applying such cdf to the data results in a uniform distributed PIT then $F$ is a valid prediction. If not, $F$ is not the suited cdf for the considered data. Figure \ref{fig:pit} provides an example, applying the true cdf results in a well calibrated PIT (left). Alternatively, applying a bad cdf results in either a overdispersed (middle) or underdispersed (right) PIT.

### 2024-04-29 thesis/input_files/point.tex
April 28 2024
correct names in contents,list of fig and tables

\section{Autoregressive models}
% explain their theory
% and how the procedure how they are used
% IMPORTANT: load series is not a stationary series so before applying AR we have to perform stationary tests or differencing steps
Autoregressive models are a standard approach for modelling time series data.
Before introducing the popular model in EPF, it is important to remember that autoregressive models assume the time series to be stationary. On the other hand, load and price have been observed to be a non stationary time series. 
Basically, stationarity means that the distribution of any subsequence of random variables of the stochatic process is invariant to shifts along the time dimension.
\\
Therefore, checking stationarity is a crucial step in applying this class of models. %(unit root tests)
Furthermore, should the time series be non stationary, we have to convert it to a stationary one by a combination of either differencing, detrending and/or log transforming it.
\subsection{ARIMA}
ARIMA(p,d,q) models are of the form
\begin{equation}
    \hat{L}_N=\sum\limits_{i=1}^{p}\phi_i L_{N-i}^{(d)}+\sum\limits_{j=1}^{q}\theta_j \epsilon_{N-j}+\epsilon_N
\end{equation}
where the parameter p is the order of the AR part, d is the degree of integrated differencing and q is the order of the MA model. $\epsilon_{N-j}$ are the observed errors in the past while the differenced term $L_N^{(d)}$ is defined recursively as $L_N^{(d)}=L_N^{(d-1)}-L_{N-1}^{(d-1)}$.
ARIMA model are compactly written as
\begin{equation}
    \phi(B)\nabla^d y_t=\theta(B)\epsilon_t
\end{equation}
where $B$ is the backward shift operator, $B^h y_t=y_{t-h}$ and $\phi(B)$ is shorthand for $\phi(B)=1-\phi_1B-\dots-\phi_p B^p$.
Similarly, $\theta(B)=1+\theta_1B+\dots+\theta_q B^q$ and $\nabla_h$ is the lag h differencing operator $\nabla_h y_t=(1-B^h)y_t=y_t-y_{t-h}$
\\
The most popular procedure for estimating this class of models is the Box-Jenkins method \cite{box2015time}.
The order p is identified by taking the lags at which the sample partial autocorrelation function (PACF) falls outside its 95\% confidence interval.
In the same way, the order q is estimated by considering the sample autocorrelation function of the observed time series. Finally, we compare ARIMA models with different p,d,q (in a neighbourhood of the identified orders) and choose the ones minimising either AIC or BIC criteria. 
% Notice, models are obtained through the maximum likelihood estimation framework.

\subsection{ARIMAX}
ARIMAX model adds explanatory variables to our ARIMA, hence, we have
\begin{equation}
    \hat{L}_N=\sum\limits_{k=1}^{h} \mu_i X_{N-k}+\sum\limits_{i=1}^{p}\phi_i L_{N-i}^{(d)}+\sum\limits_{j=1}^{q}\theta_j \epsilon_{N-j}+ \epsilon_{N}
\end{equation}
\subsection{SARIMA and SARIMAX}
SARIMA and SARIMAX incorporate seasonalities into our time series models. The notation for these models is ARIMA$(p,d,q)(P,D,Q)_m$. The second term $(P,D,Q)_m$ represents the autoregressive structure of the seasonal patter, where $m$ indicates the number of seasonalities of our time series.
For instance, in modelling series of hourly data with daily periodicity, we would set m=24.
Writing SARIMA in compact notation we have
\\
\begin{equation}
    \phi(B)\Phi(B^s)\nabla^d\nabla_m^Dy_t=\theta(B)\Theta(B^s)\epsilon_t
\end{equation}

\section{Generalized additive model}
% in standard regression the predicted y are normal since error are normal. But if we want to model say a y bernoulli which is the rate of success we can link the domain of bernoulli to that of gaussian with a link function
Generalized additive models take the form
\begin{equation}
    g[\mathbb{E}(Y_i)]=\alpha+f_1(X_1)+\dots+f_p(X_p)
\end{equation}
$f_i$ are smooth functions which are fitted using cubic smoothing splines.
g is called link function and characterizes the specific GAM model \cite{hastie2017generalized}. Approximations for $f_i$ are obtained through an iterative procedure, the backfitting algorithm.

@@

\subsection{MLP}
In order to cover a richer space of models we can stack neurons, doing so we obtain the so called multi layer perceptrons (MLPs). Hyperparameters of these models are the number of hidden layers, the number of neurons in each of those layers and the kind of activation function.
Notice, depending on the numbers of layers, MLPs are sometimes referred to as deep neural networks.

\subsubsection{NARX}
Neural networks that have as inputs past values of the target variable and other independent variables are called nonlinear autoregressive exogenous models.

\subsection{LSTM}
% https://colah.github.io/posts/2015-08-Understanding-LSTMs/

% https://rtavenar.github.io/deep_book/book_en.pdf
LSTM extends RNN by introducing a cell state in its layers. In this way, the current state of a cell depndes on the current value $X_t$, on the previous cell activation and on the previous cell state $C_{t-1}$, see figure \ref{fig:lstm_cell} for a visual representation. 
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/lstm_cell.png}
    \caption{LSTM cell}
    \label{fig:lstm_cell}
\end{figure}
The cell is responsible for deciding whether to store/forget long and short term information. To do so, each cell is made of a forget gate, an input gate and an output gate.
The forget gate takes in $h_{t-1}$ and $x_t$ and outputs a real valued vector with elements in the domain $[0,1]$, this corresponds to the ratio of information retention for each element of the cell state $C_{t-1}$.
\\
Its equation is given by
\begin{equation}
    f_t=\sigma(W_f\dot[h_{t-1},x_t]+b_f)
\end{equation}
The input gate is made up of two parts. First, a sigmoid layer decides which values of the cell state will be updated. Second, a tahn layer creates a vector of candidates $\tilde{C}_t$ to add to the current state.
\begin{equation}
    \begin{aligned}
        i_t=&\sigma(W_i\dot[h_{t-1},x_t]+b_i)
        \\
        \tilde{C}_t=&tahn(W_c\dot[h_{t-1},x_t]+b_C)
    \end{aligned}
\end{equation}
The cell state is then updated by
\begin{equation}
    C_t=f_t\odot C_{t-1}+i_t\odot \tilde{C}_t
    % odot is elementwise product
\end{equation}
Finally, we output the current cell state and the tanh activated cell state filtered by the sigmoid layer on the current input and the previous hidden state.
\begin{equation}
    \begin{aligned}
    o_t=& \sigma(W_o[h_{t-1},x_t]+b_o)
    \\
    h_t=& o_t \odot tahn(C_t)
\end{aligned}
\end{equation}
% \subsection{DeepAr} deepar é probabilistic

>>>>

\section{Autoregressive models}
% explain their theory
% and how the procedure how they are used
% IMPORTANT: load series is not a stationary series so before applying AR we have to perform stationary tests or differencing steps
Autoregressive models are a standard approach for modelling time series data.
Before introducing the popular model in EPF, it is important to remember that autoregressive models assume the time series to be stationary. On the other hand, load and price have been observed to be a non stationary time series. 
Basically, stationarity means that the distribution of any subsequence of random variables of the stochatic process is invariant to shifts along the time dimension.
\\
Therefore, checking stationarity is a crucial step in applying this class of models. %(unit root tests)
Furthermore, should the time series be non stationary, we have to convert it to a stationary one by a combination of either differencing, detrending and/or log transforming it.
\subsection{Autoregressive integrated moving average}
ARIMA(p,d,q) models are of the form
\begin{equation}
    \hat{L}_N=\sum\limits_{i=1}^{p}\phi_i L_{N-i}^{(d)}+\sum\limits_{j=1}^{q}\theta_j \epsilon_{N-j}+\epsilon_N
\end{equation}
where the parameter p is the order of the AR part, d is the degree of integrated differencing and q is the order of the MA model. $\epsilon_{N-j}$ are the observed errors in the past while the differenced term $L_N^{(d)}$ is defined recursively as $L_N^{(d)}=L_N^{(d-1)}-L_{N-1}^{(d-1)}$.
ARIMA model are compactly written as
\begin{equation}
    \phi(B)\nabla^d y_t=\theta(B)\epsilon_t
\end{equation}
where $B$ is the backward shift operator, $B^h y_t=y_{t-h}$ and $\phi(B)$ is shorthand for $\phi(B)=1-\phi_1B-\dots-\phi_p B^p$.
Similarly, $\theta(B)=1+\theta_1B+\dots+\theta_q B^q$ and $\nabla_h$ is the lag h differencing operator $\nabla_h y_t=(1-B^h)y_t=y_t-y_{t-h}$
\\
The most popular procedure for estimating this class of models is the Box-Jenkins method \cite{box2015time}.
The order p is identified by taking the lags at which the sample partial autocorrelation function (PACF) falls outside its 95\% confidence interval.
In the same way, the order q is estimated by considering the sample autocorrelation function of the observed time series. Finally, we compare ARIMA models with different p,d,q (in a neighbourhood of the identified orders) and choose the ones minimising either AIC or BIC criteria. 
% Notice, models are obtained through the maximum likelihood estimation framework.

\subsection{Autoregressive integrated moving average with exogenous variables}
ARIMAX model adds explanatory variables to our ARIMA, hence, we have
\begin{equation}
    \hat{L}_N=\sum\limits_{k=1}^{h} \mu_i X_{N-k}+\sum\limits_{i=1}^{p}\phi_i L_{N-i}^{(d)}+\sum\limits_{j=1}^{q}\theta_j \epsilon_{N-j}+ \epsilon_{N}
\end{equation}
\subsection{Seasonal autoregressive integrated moving average and Seasonal autoregressive integrated moving average with exogenous variables}
SARIMA and SARIMAX incorporate seasonalities into our time series models. The notation for these models is ARIMA$(p,d,q)(P,D,Q)_m$. The second term $(P,D,Q)_m$ represents the autoregressive structure of the seasonal patter, where $m$ indicates the number of seasonalities of our time series.
For instance, in modelling series of hourly data with daily periodicity, we would set m=24.
Writing SARIMA in compact notation we have
\\
\begin{equation}
    \phi(B)\Phi(B^s)\nabla^d\nabla_m^Dy_t=\theta(B)\Theta(B^s)\epsilon_t
\end{equation}

\section{Generalized additive model}
% in standard regression the predicted y are normal since error are normal. But if we want to model say a y bernoulli which is the rate of success we can link the domain of bernoulli to that of gaussian with a link function
Generalized additive models (GAMs) take the form
\begin{equation}
    g[\mathbb{E}(Y_i)]=\alpha+f_1(X_1)+\dots+f_p(X_p)
\end{equation}
$f_i$ are smooth functions which are fitted using cubic smoothing splines.
g is called link function and characterizes the specific GAM model \cite{hastie2017generalized}. Approximations for $f_i$ are obtained through an iterative procedure, the backfitting algorithm.

@@

\subsection{Multilayer perceptrons}
In order to cover a richer space of models we can stack neurons, doing so we obtain the so called multi layer perceptrons (MLPs). Hyperparameters of these models are the number of hidden layers, the number of neurons in each of those layers and the kind of activation function.
Notice, depending on the numbers of layers, MLPs are sometimes referred to as deep neural networks.

% \subsubsection{NARX}
% Neural networks that have as inputs past values of the target variable and other independent variables are called nonlinear autoregressive exogenous models.

\subsection{Long short term memory}
% https://colah.github.io/posts/2015-08-Understanding-LSTMs/

% https://rtavenar.github.io/deep_book/book_en.pdf
LSTM extends RNN by introducing a cell state in its layers. In this way, the current state of a cell depndes on the current value $X_t$, on the previous cell activation and on the previous cell state $C_{t-1}$, see figure \ref{fig:lstm_cell} for a visual representation. 
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/lstm_cell.png}
    \caption{Long short term memory cell}
    \label{fig:lstm_cell}
\end{figure}
The cell is responsible for deciding whether to store/forget long and short term information. To do so, each cell is made of a forget gate, an input gate and an output gate.
The forget gate takes in $h_{t-1}$ and $x_t$ and outputs a real valued vector with elements in the domain $[0,1]$, this corresponds to the ratio of information retention for each element of the cell state $C_{t-1}$.
\\
Its equation is given by
\begin{equation}
    f_t=\sigma(W_f\dot[h_{t-1},x_t]+b_f)
\end{equation}
The input gate is made up of two parts. First, a sigmoid layer decides which values of the cell state will be updated. Second, a tahn layer creates a vector of candidates $\tilde{C}_t$ to add to the current state.
\begin{equation}
    \begin{aligned}
        i_t=&\sigma(W_i\dot[h_{t-1},x_t]+b_i)
        \\
        \tilde{C}_t=&tahn(W_c\dot[h_{t-1},x_t]+b_C)
    \end{aligned}
\end{equation}
The cell state is then updated by
\begin{equation}
    C_t=f_t\odot C_{t-1}+i_t\odot \tilde{C}_t
    % odot is elementwise product
\end{equation}
Finally, we output the current cell state and the tanh activated cell state filtered by the sigmoid layer on the current input and the previous hidden state.
\begin{equation}
    \begin{aligned}
    o_t=& \sigma(W_o[h_{t-1},x_t]+b_o)
    \\
    h_t=& o_t \odot tahn(C_t)
\end{aligned}
\end{equation}
% \subsection{DeepAr} deepar é probabilistic

### 2024-04-29 thesis/input_files/problemdescription.tex
correct names in contents,list of fig and tables

Individuals and organizations constantly face situations of uncertainty thus the need of robust forecasting methods. Such methods are crucial to the process of taking informed decision and to strategic planning.
\\
The basic idea of forecasting is that we can extract knowledge from the past in order to make educated guesses about the future. Consequently, the range of fields where forecasting can be applied is very wide.
In this thesis, our focus lies on applying forecasting techniques to the energy sector. 
\\
The reason of our decision to focus on the energy market is mainly motivated by the rapid changes it has undergone. %/experienced 
Over the last decades, electricity markets have gone through an unprecedented transformation; this shift was driven by the liberalization of such markets, the development and integration of renewable energy sources, the  increase of low carbon technologies and the adoption of smart meters. Events like the California electricity crisis help also motivating the choice of the electricity sector as subject of our studies, see \cite{california}.
Furthermore, the process of deregulation lead to an increasing interest in the field of electricity price forecasting (EF) within the academic community \ref{fig:epf_evolution}.
\\
In addtion, the United Nations have identified the right to access affordable, reliable, sustainable and modern energy as one of their 17 SDGs \cite{un_sdgs}.
\\
Finally, the electricity market has a set of features that make it unique: electricity cannot be stored in an efficient way and supply and demand have to be matched instantly.
\\
\section{Motivation}
The are mutliple reasons why the energy sector needs robust forecasting techniques.
For power market companies, being able to predict prices with a low MAPE \ref{mape} results in increased savings \cite{savings}. Furthermore,the adoption of smart meters provides power market companies with a ton of consumer data; this can enable them to better model consumer preferences.
\\
Transmission system operators' main goal is to match supply and deman, generally TSO do so by increasing or decreasing the generation supplied. Thus, from their view point forecasting is critical for balancing the electricity netwrok.
\\
Probabilistic forecasting may be useful to power producers, traders and consumers in order to improve their decision making process and managing risk(VaR). This holds in particular for traders, because probabilistic forecast enables them to simulate scenarios and carry out stress tests.
\\
Other possible applications are: control of storage, demand side response, anomaly detection, network design and planning, simulating inputs and handling missing data.
\\
\section{Point vs probabilistic forecast}
A distinction has to be made between two types of forecasting approaches: point forecasts and probabilistic forecasts.
Point prediction, also called deterministic forecasting in the literature, is all about predicting a particular value in time.
On the other hand, with probabilistic forecasting we aim at predicting either a prediction interval, quantiles or a probability distribution for each point in time. For this reason, probabilistic forecasts are more informative than point forecasts; and this is why the interest of the research community is shifting towards them.
Note that a probabilistic forecast can be turned into a point forecast by simply taking its expecatation.
Alternatively, a probabilistic forecast can be derived from a point one by modeling the residuals of the point prediction.

>>>>

Individuals and organizations constantly face situations of uncertainty thus the need of robust forecasting methods. Such methods are crucial to the process of taking informed decision and to strategic planning.
\\
The basic idea of forecasting is that we can extract knowledge from the past in order to make educated guesses about the future. Consequently, the range of fields where forecasting can be applied is very wide.
In this thesis, our focus lies on applying forecasting techniques to the energy sector. 
\\
The reason of our decision to focus on the energy market is mainly motivated by the rapid changes it has undergone. %/experienced 
Over the last decades, electricity markets have gone through an unprecedented transformation; this shift was driven by the liberalization of such markets, the development and integration of renewable energy sources, the  increase of low carbon technologies and the adoption of smart meters. Events like the California electricity crisis help also motivating the choice of the electricity sector as subject of our studies, see \cite{california}.
Furthermore, the process of deregulation lead to an increasing interest in the field of electricity price forecasting (EF) within the academic community \ref{fig:epf_evolution}.
\\
In addtion, the United Nations have identified the right to access affordable, reliable, sustainable and modern energy as one of their 17 SDGs \cite{un_sdgs}.
\\
Finally, the electricity market has a set of features that make it unique: electricity cannot be stored in an efficient way and supply and demand have to be matched instantly.
\\
\section{Motivation}
The are mutliple reasons why the energy sector needs robust forecasting techniques.
For power market companies, being able to predict prices with a low MAPE \ref{mape} results in increased savings \cite{savings}. Furthermore,the adoption of smart meters provides power market companies with a ton of consumer data; this can enable them to better model consumer preferences.
\\
Transmission system operators' main goal is to match supply and deman, generally TSO do so by increasing or decreasing the generation supplied. Thus, from their view point forecasting is critical for balancing the electricity netwrok.
\\
Probabilistic forecasting may be useful to power producers, traders and consumers in order to improve their decision making process and managing risk(VaR). This holds in particular for traders, because probabilistic forecast enables them to simulate scenarios and carry out stress tests.
\\
Other possible applications are: control of storage, demand side response, anomaly detection, network design and planning, simulating inputs and handling missing data.
\\
\section{Point versus probabilistic forecast}
A distinction has to be made between two types of forecasting approaches: point forecasts and probabilistic forecasts.
Point prediction, also called deterministic forecasting in the literature, is all about predicting a particular value in time.
On the other hand, with probabilistic forecasting we aim at predicting either a prediction interval, quantiles or a probability distribution for each point in time. For this reason, probabilistic forecasts are more informative than point forecasts; and this is why the interest of the research community is shifting towards them.
Note that a probabilistic forecast can be turned into a point forecast by simply taking its expecatation.
Alternatively, a probabilistic forecast can be derived from a point one by modeling the residuals of the point prediction.
